Partial rank data with the \pkg{hyper2} package: likelihood functions for generalized Bradley-Terry models

### 2019-05-03 inst/collusion_chess_draws.Rnw
use Rnw format, as I need sophisticated tables not available in Rmd format

---
title: "Collusion at the 1962 chess world championships"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{collusion at the 1962 world chess championships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

First load the package:
```{r}
library("hyper2")
options("width" = 90)
```

@@

\[
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\]

@@

\[
\prod_{i<j}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\]

@@

\[
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
\]

@@

```{r}
jj <- read.table("stockholm1962.txt",header=FALSE)
results <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(results) <- players
colnames(results) <- substr(players,1,1)
results
```

@@

```{r}
all(t(results) == 2-results)
which(results != 2-t(results),arr.ind=TRUE)
```

@@

```{r}
H <- hyper2(pnames=c("draw","collusion",players))

for(i in seq_len(nrow(results)-1)){  # i=row
    for(j in seq(from=i+1,to=ncol(results))){ # j = col
        o <- results[i,j]
        if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
            monster <- "collusion"
        } else {
            monster <- "draw" 
        }
        if(o==2){ # row player wins
            H[players[i]] %<>% inc(1)
        } else if(o==0){  # col player wins
            H[players[j]] %<>% inc(1)
        } else if(o==1){ # game drawn
            H[monster] %<>% inc(1)
        } else {
            stop("this cannot happen")
        }
        H[c(players[i],players[j],monster)] %<>% dec(1)
    }  # j loop closes
} # i loop closes
```
and we have

```{r}
head(H)
dotchart(maxp(H),pch=16)
dotchart(log(maxp(H)),pch=16)
```

which would indicate strongly that collusion is indeed a strong
feature of the tournament.  Now some optimization.  First optimize freely:

```{r}
max_support_free <- maxp(H,give=TRUE)$value
ml_p   <- maxp(H)

s <- indep(equalp(H))
small <- 1e-3
s[1] <- s[1] + small
s[2] <- s[2] - small
```
Perform the constrained optimization:
```{r}
ml_p_constrained <- maxp(H,fcm=c(1,-1,rep(0,22)),fcv=0,startp=s)
max_support_constrained <- loglik(H,indep(ml_p_constrained))
max_support_constrained
```

Now compare the supports:

```{r}
extra_support <- max_support_free - max_support_constrained
extra_support
```

well in excess of Edwards's two-unit criterion.  Alternatively, we may
apply Wilks's theorem and observe that the test statistic is in the
tail region of its asymptotic distribution:

```{r}
pchisq(2*extra_support,df=1,lower.tail=FALSE)
```

# Now Curacao

This is the 5th candidates Curacao 1962 results table:

```{r}
jj <- read.table("curacao1962.txt",header=FALSE)
curacao <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(curacao) <- players
colnames(curacao) <- players
curacao[is.na(curacao)] <- ""
curacao <- noquote(curacao)
curacao
```

Now convert to a likelihood function:
```{r}
Hcur <- hyper2(pnames=c("draw","collusion",players))
for(i in seq_len(nrow(curacao)-1)){  # i=row
  for(j in seq(from=i+1,to=ncol(curacao))){ # j = col
    r <- strsplit(curacao[i,j],'')[[1]]
    if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
      monster <- "collusion"
    } else {
      monster <- "draw"
    }
    if(any(r=="W")){  # row player wins
      wins <- sum(r=="W")
      Hcur[  players[i]                    ] %<>% inc(wins)
      Hcur[c(players[i],players[j],monster)] %<>% dec(wins)
    }
    if(any(r=="L")){  # row player loses
      losses <- sum(r=="L")
      Hcur[             players[j]         ] %<>% inc(losses)
      Hcur[c(players[i],players[j],monster)] %<>% dec(losses)
    }
    if(any(r=="W")){  # row player draws
      draws <- sum(r=="D")
      Hcur[                        monster ] %<>% inc(draws)
      Hcur[c(players[i],players[j],monster)] %<>% dec(draws)
    }
  }  # j loop closes
} # i loop closes
Hcur
```

```{r}
dotchart(maxp(Hcur))
```

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% just as usual
\author{Robin K. S. Hankin\\Auckland University of Technology}
\title{Partial rank data with the \pkg{hyper2} package: likelihood
  functions for generalized Bradley-Terry models}

%\VignetteIndexEntry{The hyper2 package}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Robin K. S. Hankin}
\Plaintitle{Chess Draws}
\Shorttitle{Chess Draws}

%% an abstract and keywords
\Abstract{

Here I analyse chess games for draws using the hyper2 package.
}

\Keywords{Dirichlet distribution, hyperdirichlet distribution,
  combinatorics, \proglang{R}, multinomial distribution, constrained
optimization, Bradley-Terry}
\Plainkeywords{Dirichlet distribution, hyperdirichlet distribution,
  combinatorics, R, multinomial distribution, constrained
optimization, Bradley-Terry}

%% publication information
%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
%% If it was not (yet) accepted, leave them commented.
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  Auckland University of Technology\\
  E-mail: \email{hankin.robin@gmail.com}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\bolda}{\mathbf a}
\newcommand{\boldp}{\mathbf p}
\newcommand{\boldV}{\mathbf V}
\newcommand{\boldalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\boldzero}{\mathbf 0}

\newcommand{\mcf}{\mathcal F}
\newcommand{\mcm}{\mathcal M}
\newcommand{\mcs}{\mathcal S}
\newcommand{\mcw}{\mathcal W}

\newcommand{\pG}{p_{\mbox{\frakfamily g}}}

\SweaveOpts{}
\begin{document}
\section{Introduction}

<<echo=FALSE,print=FALSE>>=
ignore <- require(magrittr,quietly=TRUE)
ignore <- require(hyper2,quietly=TRUE)
options("width" = 90)
@

\section{Introduction: the Bradley-Terry model}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\end{equation}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\end{equation}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
\end{equation}

@@

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays white, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  13 &   2  \\
  Kr &  5 &  -  &   7  \\
  Ks & 11 &   9 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between three
  Grandmasters.  Thus, for example, Anand played white against Karpov
  a total of $18+20+13=51$ times and won 18, drew 20, lost 13}
\end{table}

The likelihood function is

<<>>=
jj <- read.table("stockholm1962.txt",header=FALSE)
results <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(results) <- players
colnames(results) <- substr(players,1,1)
results
@

@@

<<>>=

all(t(results) == 2-results)
which(results != 2-t(results),arr.ind=TRUE)
@

@@

\bibliography{hyper2}
\end{document}
